Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}

### 2022-09-13 paper/paper.tex
edits

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
  pdfauthor={Dean Marchiori},
  pdfkeywords={NEDMABA, SMPD4},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Clinical Phenotype and Survival Analysis in Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}}
\author[]{Dean Marchiori%
\corref{cor1}%
}
 \ead{deanmarchiori@gmail.com}

@@

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This research aims to
collate and standardise available case reports into a common dataset, to
analyse and identify meaningful clusters in the clinical phenotype, and
to quantify the survival probability for children with NEDMABA. To
overcome the challenge of multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To account for censoring in the data Kaplan-Meier estimation is
formulated calculate patient survival time. The analysis correctly
detected the classic phenotype for this condition, as well as an
additional distinct feature-cluster relating to findings of vocal cord
paralysis, feeding dysfunction and respiratory failure. Several
additional outlying and isolated clinical features were also identified.
The survival probability for those affected was found to decline sharply
in early infancy with median survival of 150 days, but with some
surviving as long as 12.5 years. This wide range of outcomes is
provisionally associated with different variant types however this
conclusion could not be validated based on very low sample sizes. This
analysis represents the first of its kind to help describe associations
and trajectories of individuals with this newly reported condition,
despite challenges with sparse and inconsistent data. This analysis can
provide clinicians and genetic counsellors with better information to
aide in decision making and support for families with this rare
condition.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, enhanced, boxrule=0pt, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

\citet{ravenscroft2021neurogenetic} performed a study over 190 probands
and identified a novel missense variation in SMPD4 which at the time of
whole exome sequencing was not well described in the literature. The
impacted family from Melbourne exhibited features involving
arthrogryposis multiplex congenita, complex brain malformations, small
for gestation age and hypoplasia of the corpus callosum. In two of the
three related cases were additional features of microcephaly, congenital
encephalopathy, cerebellar malformation and hypoplasia and
hypomyelination.

A detailed study by \citet{magini2019loss} involved 12 unrelated
families with 32 individuals (21 with detailed clinical information).
Presentations were of microcephaly, simplified gyral pattern of the
cortex, hypomyelination, cerebellar hypoplasia, congenital
arthrogryposis, and early fetal/postnatal demise. Despite this being the
largest cohort studied, the clinical features and survival times among
participants varied greatly. Three missense changes were noted in the
study with affected children in these families often showing a milder
presentation suggestive of possible residual function. In these cases
individuals were able to develop independent motor skills, have mild
intellectual disability and arthrogryposis without evidence of
simplified gyral patterns on brain MRI. Other patients with truncating
variants are shown to have more severe presentations, while a range of
additional significant clinical features were reported involving
dysmorphic facial features, seizure, vocal cord paralysis and hearing
impairment.

A case study from China is described in \citet{ji2022case} involving a
girl presenting in infancy with intrauterine growth restriction,
microcephaly, postnatal developmental delay, arthrogryposis,
hypertonicity, seizure, and hypomyelination on brain magnetic resonance
imaging. The authors report parallels with two cases showing the same
homozygous null variant. An individual reported in
\citet{monies2019lessons} presented with distinct symptoms of brain
atrophy and skeletal dysplasia whereas the case in
\citet{magini2019loss} with the same variant exhibited more typical
clinical features.

A recent study by \citet{bijarnia2022growth} presents the case of a
22-month old girl presenting with the typical phenotype of
neurodevelopmental delay, prenatal onset growth failure, arthrogryposis,
microcephaly and brain anomalies including severe hypomyelination,
simplified gyral pattern and hypoplasia of corpus callosum and
brainstem. Notably, there was also additional non-typical clinical
findings of nystagmus and visual impairment secondary to macular
dystrophy and retinal pigment epithelial stippling at posterior pole.

Further work to collate and analyse data relating to NEDMABA is
challenging. While \citet{magini2019loss} cataloged a detailed clinical
phenotype data set as a supplementary data set to their study, these
data are not suitable for statistical analysis as presented. Many of the
clinical features are represented as free text descriptions and a
variety of non-standard terminologies are applied between cases, making
machine interpretation of the data difficult. Other studies
\citep{ji2022case, bijarnia2022growth} present only written case reports
with some tabulated summaries. While larger studies
\citep{ravenscroft2021neurogenetic, monies2019lessons} focus more on
genetic data, with less detail included on the clinical presentation of
individuals in the cohort.

To overcome these data challenges, \citet{JSSv059i10} proposes a data
structure know as `tidy' data which organises each observation in a row,
with each feature as its own column and each value in just one cell. In
this case detailed text-based data can be transformed to high
dimensional binary indicators of key clinical features. This structure
is conducive to effective statistical analysis and integrates
deliberately with the \emph{tidyverse} \citep{tidyverse} collection of
packages for data analysis in R \citep{rbase}.

Further analysis to better describe rare genetic conditions was explored
in \citet{diaz2020phenotype} in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) frequently present with varied
symptoms that do not match exactly with the described phenotype. When
considering this situation of low sample size data with many features,
methods in the field of multivariate statistical analysis are commonly
deployed. Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis \citep{le2010multiple} are appropriate for high
dimensional data comprised of binary indicators. This technique is cited
in many studies in the application of dimension reduction and clustering
of comorbidities and phenotypes from disease
\citep{han2018cluster, costa2013use}.

@@

An open research question exists around the clinical pathway and
survival time of children exhibiting bi-allelic loss-of-function
variants in \emph{SMPD4}. With current research highlighting a diverse
and heterogeneous phenotype, the relationship and correlations between
these diverse features is not well understood. Furthermore, the survival
time of affected children has not been extensively analysed despite
early-demise featuring as a severe outcome in many cases. While a
connection has been highlighted between some missense variants and a
milder presentation with longer survival \citep{magini2019loss}, this
hypothesis has not been analysed further.

This research has three key aims. First, to collate and transform the
variety of early case reports and studies on clinical phenotype data for
this novel variant into an analysis-ready \emph{tidy} data set.
Secondly, to conduct analysis on the associations between commonly
reported clinical features. Finally, to statistically quantify the
expected survival time of children with NEDMABA based on the current
case reports.

@@

The data were read into \texttt{R} \citep{rbase} without any changes, so
as to preserve the reproducibility of the data transformation steps. The
data were transposed to ensure it could be presented as \emph{tidy}
formatted data \citep{JSSv059i10}. This requires each variable (clinical
phenotype) forms a column; each observation (individual) forms a row and
each type of observational unit forms a table (every cell has just one
item). This resulted in a long-formatted data set of 21 rows
(individuals) and 64 columns (clinical features). In many cases, key
clinical information was entered as free-text descriptions, which
rendered any attempt of meaningful analysis impractical
(Table~\ref{tbl-nontidy}). In these cases, the text was tokenised by
separating the list of clinical observations at each comma and forming a
binary indicator column noting its presence `1' or absence `0'
(Table~\ref{tbl-tidy}).

@@

The final data set consisted of 28 observations (one per individual) and
152 variables (one per clinical feature). These variables were comprised
of qualitative text descriptions, binary indicators of features as
described above and other fields including background information on the
patient and various bio markers. A full list of this data set is
available in Table~\ref{tbl-dict}. These data sets were compiled into a
publicly available \texttt{R} Package called \texttt{SMPD4}
\citep{smpd4-data} in order to allow for reproducibility and sharing.

@@

Cluster validation measures resulted in either 2 or 5 clusters as the
optimal choice (Table~\ref{tbl-opticlust}). In order to partition the
data into meaningful groupings, k-means with 5 clusters was selected.
This reflects the optimal choice for both AD and FOM cluster metrics. A
scatter plot of the cluster results shown in Figure~\ref{fig-clust} is
projected on the first two principal dimensions highlights a core
cluster of classic features of NEDMABA such as microcephaly,
hypomyelination, arthrogryposis and feet deformity and structural brain
anomalies. A separate and adjacent cluster is formed of airway and
feeding related conditions involving vocal cord paralysis, tracheostomy,
tube feeding and enlarged ventricles. Three other smaller satellite
clusters exists where highly distinctive features were associated with a
single individual or family. A full articulation of feature to cluster
assignment is given in Table~\ref{tbl-profile}.

\hypertarget{tbl-opticlust}{}
\begin{longtable}{lrlr}
\caption{\label{tbl-opticlust}Optimal clusters }\tabularnewline

@@

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562
days) with median survival time calculated as 150 days. At time 0 the
survival probability was 0.82 (95\% CI: 0.69, 0.98). Survival
probability after one year at 0.403 (95\% CI: 0.252, 0.644) and survival
probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52).

@@

\includegraphics{paper_files/figure-pdf/unnamed-chunk-14-1.pdf}

@@

Often, data collected in studies is fit for descriptive purposes but not
for in-depth statistical analysis. This meta-review of current studies
was able to consolidate and make publicly available, data sets that
allow analysis to be conducted on emerging conditions such as NEDMABA.

The use of dimension reduction techniques were shown to be successful in
uncovering relevant and novel substructure in the data. This was
particularly evident with the detection of the accepted typical
phenotype of microcephaly, hypomyelination, arthrogryposis, simplified
gyral pattern and developmental delay.

After excluding two outlying remarkable cases, this analysis clearly
identifies the presence of an adjacent cluster of related features that
are not as well described in The Human Phenotype Ontology
\citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. We identify the
prolific feature of `respiratory distress' as an expression more
commonly associated with \emph{other} features such as vocal cord palsy,
tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
These features are present in the case reported data, but not well
described in HPO and help account for the variation seen in the clinical
phenotype in the first few principal dimensions.

Other smaller clusters identify outlying features that are expressed in
individuals or specific families. Many of the outlying features are
highly specific facial or anatomical dysmorphisms. While coarse facial
features are a noted and core presentation, those with a milder course
had clinical reports that articulated these dysmorphisms in much more
detail, which skewed the interpretation.

On a similar note, talipes appears as an isolated feature without the
typical microcephaly and hypomyelination. However, there was significant
heterogeneity in how feet contractures were reported in clinical case
reports. While distinctions were made during data pre-processing between
rocker-bottom feed and talipes, this likely to make intpretation
somewhat difficult. This is particularly clear as rocker-bottom feet was
found closely related in our analysis to other core features despite the
prevalence in HPO being quite low (2/21), suggesting possible
under-reporting or confounded reporting of this feature.

Another feature clusted within core identified features is bilateral
cleft lip. This condition has a reported frequency of just 2/21
(Table~\ref{tbl-hpo}) and may be an emerging feature for a more typical
presentation given its close relationship with other classic features.

The heterogeneous variation in the data was clearly evident however,
with the first four principal axes being required to explain more than
50\% of the variation in the data set. Visual inspection of the data
projected into this lower dimensional space indicate the primary sources
of variation in the clinical phenotype are explained by a few
individuals with very different clinical features. This has confounded
the typical or classical presentation. Not all features in the data were
included in this analysis, particular if isolated to just one case. In
this circumstance, these terms in Table~\ref{tbl-hpo} have been labelled
as `Isolated'.

While the number of clusters chosen all represent valid partition of the
data, some metrics indicated an optimal selection of 2 clusters. This
type of partition is useful when segregating outliers from the rest of
the data and is valid in some applications. In this study, a larger
number (5) was selected to align the goals of the analysis which was to
uncover distinct groups of features.

In terms of patient survival, the first few months show a sharp decline
in survival probability confirming early demise in the infant period as
a key factor. A small number of cases show much longer survival into
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). On
the other hand, a naive estimate using just the number of deceased
patients at this time as proportion of the study population is 0.57
(16/28). This highlights the importance of applying a survival analysis
technique that incorporates censoring so as not to erroneously
overestimate the survival probability. The current literature suggests
the type of gene variant has an impact on survival time, which seems to
be supported in this work. Through a comparison of multiple survival
curves, stratified by variation type, the limited sample size prevented
definitive conclusions with many cells having expected counts less than
5.

>>>>

\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{longtable}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in SMPD4},
  pdfauthor={Dean Marchiori},
  pdfkeywords={NEDMABA, SMPD4},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{frontmatter}
\title{Clinical Phenotype and Survival Analysis of Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies Due to Bi-allelic Loss of Function Variants in \emph{SMPD4}}
\author[]{Dean Marchiori%
\corref{cor1}%
}
 \ead{deanmarchiori@gmail.com}

@@

\begin{abstract}
A recently described, rare genetic condition known as Neurodevelopmental
Disorder with Microcephaly, Arthrogryposis, and Structural Brain
Anomalies (NEDMABA) has been identified in children with bi-allelic
loss-of-function variants in \emph{SMPD4}. The progression of this
condition is not well understood with the limited case reports described
so far exhibiting a severe and clinically diverse phenotype. A gap
exists in the understanding of associations present in the heterogenous
features of the clinical phenotype, and the expected survival
probabilities of affected individuals. This is driven in part to the
paucity of analysis-ready data on reported cases. This analysis aims to
collate and standardise available case reports into a common dataset, to
analyse and identify meaningful clusters in the clinical phenotype, and
to quantify the survival probability for children with NEDMABA. To
overcome the challenge of multidimensional data on very few subjects, we
employ Multiple Correspondence Analysis (MCA) as a dimension reduction
technique, which is then subject to cluster analysis and interpretation.
To account for censoring in the data, Kaplan-Meier estimation is
formulated to calculate patient survival time. The analysis correctly
detected the classic phenotype for this condition, as well as a new
distinct feature-cluster relating to findings of vocal cord paralysis,
feeding dysfunction and respiratory failure. The survival probability
for those affected was found to decline sharply in early infancy with
median survival of 150 days, but with some surviving as long as 12.5
years. This wide range of outcomes is provisionally associated with
different variant types however this conclusion could not be validated
based on very low sample sizes. An \texttt{R} package called
\texttt{SMPD4} was developed to publish standardised analysis-ready
datasets used in this study. This analysis represents the first of its
kind to help describe associations and trajectories of individuals with
this newly reported condition, despite challenges with sparse and
inconsistent data. This analysis can provide clinicians and genetic
counsellors with better information to aide in decision making and
support for families with this rare condition.
\end{abstract}

\begin{keyword}
    NEDMABA \sep 
    SMPD4
\end{keyword}
\end{frontmatter}\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt, enhanced, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

A detailed study by \citet{magini2019loss} involved 12 unrelated
families with 32 individuals (21 with detailed clinical information).
Presentations were of microcephaly, simplified gyral pattern of the
cortex, hypomyelination, cerebellar hypoplasia, congenital
arthrogryposis, and early fetal/postnatal demise. Despite this being the
largest cohort studied, the clinical features and survival times among
participants varied greatly. Three missense changes were noted in the
study with affected children in these families often showing a milder
presentation suggestive of possible residual function. In these cases
individuals were able to develop independent motor skills, have mild
intellectual disability and arthrogryposis without evidence of
simplified gyral patterns on brain MRI. Other patients with truncating
variants are shown to have more severe presentations, while a range of
additional significant clinical features were reported involving
dysmorphic facial features, seizure, vocal cord paralysis and hearing
impairment. \citet{ravenscroft2021neurogenetic} studied a family from
Melbourne exhibiting features involving arthrogryposis multiplex
congenita, complex brain malformations, small for gestation age and
hypoplasia of the corpus callosum. In two of the three related cases
were additional features of microcephaly, congenital encephalopathy,
cerebellar malformation and hypoplasia and hypomyelination. A case study
from China is described in \citet{ji2022case} involving a girl
presenting in infancy with intrauterine growth restriction,
microcephaly, postnatal developmental delay, arthrogryposis,
hypertonicity, seizure, and hypomyelination on brain magnetic resonance
imaging. The authors report parallels with two cases showing the same
homozygous null variant. An individual reported in
\citet{monies2019lessons} presented with distinct symptoms of brain
atrophy and skeletal dysplasia whereas the case in
\citet{magini2019loss} with the same variant exhibited more typical
clinical features. A recent study by \citet{bijarnia2022growth} presents
the case of a 22-month old girl presenting with the typical phenotype of
neurodevelopmental delay, prenatal onset growth failure, arthrogryposis,
microcephaly and brain anomalies including severe hypomyelination,
simplified gyral pattern and hypoplasia of corpus callosum and
brainstem. Notably, there was also additional non-typical clinical
findings of nystagmus and visual impairment secondary to macular
dystrophy and retinal pigment epithelial stippling at posterior pole.

Further work to collate and analyse data relating to NEDMABA is
challenging. While \citet{magini2019loss} included a supplementary
clinical phenotype data set in their study, these data are not suitable
for statistical analysis. Many of the clinical features are represented
as text descriptions and a variety of non-standard terminologies are
applied between cases, making machine interpretation of the data
difficult. Other studies \citep{ji2022case, bijarnia2022growth} present
only written case reports with some tabulated summaries. While larger
studies \citep{ravenscroft2021neurogenetic, monies2019lessons} focus
more on genetic data, with less detail included on the clinical
presentation of individuals. To overcome these data challenges,
\citet{JSSv059i10} proposes a data structure know as `tidy' data which
organises each observation in a row, with each feature as its own column
and each value in just one cell. In this case detailed text-based data
can be transformed to high dimensional binary indicators of key clinical
features. This structure is conducive to effective statistical analysis
and integrates deliberately with the \emph{tidyverse} \citep{tidyverse}
collection of packages for data analysis in R \citep{rbase}.

Further analysis to better describe rare genetic conditions was explored
in \citet{diaz2020phenotype} in the context of large scale
genotype-phenotype analysis. The authors argue patients with rare
disorders (often in small samples) frequently present with varied
symptoms that do not match exactly with the described phenotype. When
considering the situation of low sample size data with many features,
methods in the field of multivariate statistical analysis are commonly
deployed. Here the aim is to find substructure in the data, or a simple
representation of the multi-dimensional space. Methods, such as Multiple
Correspondence Analysis \citep{le2010multiple} are appropriate for high
dimensional data comprised of binary indicators. This technique is cited
in many studies in the application of dimension reduction and clustering
of comorbidities and phenotypes from disease
\citep{han2018cluster, costa2013use}.

@@

An open research question exists around the clinical pathway and
survival time of children exhibiting bi-allelic loss-of-function
variants in \emph{SMPD4}. With current research highlighting a diverse
and heterogeneous phenotype, the relationship between these diverse
features is not well understood. Furthermore, the survival time of
affected children has not been extensively analysed despite early-demise
featuring as a severe outcome. While a connection has been highlighted
between some missense variants and a milder presentation with longer
survival \citep{magini2019loss}, this hypothesis has not been analysed
further.

This analysis has three key aims. First, to collate and transform the
variety of early case reports and studies on clinical phenotype data for
this novel variant into an analysis-ready \emph{tidy} data set.
Secondly, to conduct analysis on the associations between commonly
reported clinical features. Finally, to statistically quantify the
expected survival time of children with NEDMABA based on the current
case reports.

@@

The data were read into \texttt{R} \citep{rbase} without any changes, so
as to preserve the reproducibility of the data transformation steps. The
data were transposed to ensure it could be presented as \emph{tidy}
formatted data \citep{JSSv059i10}. This requires that each variable
(clinical phenotype) forms a column; each observation (individual) forms
a row and each type of observational unit forms a table (every cell has
just one item). This resulted in a long-formatted data set of 21 rows
(individuals) and 64 columns (clinical features). In many cases, key
clinical information was entered as free-text descriptions, which
rendered any attempt of meaningful analysis impractical
(Table~\ref{tbl-nontidy}). In these cases, the text was tokenised by
separating the list of clinical observations at each comma and forming a
binary indicator column noting its presence `1' or absence `0'
(Table~\ref{tbl-tidy}).

@@

The final data set consisted of 28 observations (one per individual) and
152 variables (one per clinical feature). These variables were comprised
of qualitative text descriptions, binary indicators of features as
described above and other fields including background information on the
patient and various bio markers. A full list of this data set is
available in Table~\ref{tbl-dict}. These data sets were compiled into a
publicly available \texttt{R} Package called \texttt{SMPD4}
\citep{smpd4-data} to allow for reproducibility and sharing.

@@

Cluster validation measures resulted in either 2 or 5 clusters as the
optimal choice (Table~\ref{tbl-opticlust}). In order to partition the
data into meaningful groupings, k-means with 5 clusters was selected.
This reflects the optimal choice for both AD and FOM cluster metrics.
The results are summarised in Table~\ref{tbl-clusttab} and a scatter
plot of the cluster results is shown in Figure~\ref{fig-clust} is
projected on the first two principal dimensions. This highlights a core
cluster of classic features of NEDMABA such as microcephaly,
hypomyelination, arthrogryposis and feet deformity and structural brain
anomalies. A separate and adjacent cluster is formed of airway and
feeding related conditions involving vocal cord paralysis, tracheostomy,
tube feeding and enlarged ventricles. Three other smaller satellite
clusters exists where highly distinctive features were associated with a
single individual or family. A full articulation of feature to cluster
assignment is given in Table~\ref{tbl-profile}.

\hypertarget{tbl-opticlust}{}
\begin{longtable}{lrlr}
\caption{\label{tbl-opticlust}Optimal cluster numbers and methods based on a range of stability
validation metrics. }\tabularnewline

@@

\hypertarget{tbl-clusttab}{}
\setlength{\LTpost}{0mm}
\begin{longtable}{rrl}
\caption{\label{tbl-clusttab}A summary table to cluster membership showing the cluster number and
number of features in each cluster and a sample of features that have
membership in each. }\tabularnewline

\toprule
Cluster No. & No. Features & Sample Features in Cluster\textsuperscript{1} \\ 
\midrule
1 & 7 & polyhydramnios, short palpebral fissures, patent ductus arteriosus, thin lips \\ 
2 & 11 & vocal cord palsy, respiratory distress, tracheostomy, posteriorly angulated ears \\ 
3 & 7 & epicanthal folds, receding forehead, spastic tetraparesis, intubation \\ 
4 & 21 & microcephaly, sgp, iugr, talipes \\ 
5 & 6 & hypercapnia, inspiratory stridor, oligohydramnios, prominent ears \\ 
\bottomrule
\end{longtable}
\begin{minipage}{\linewidth}
\textsuperscript{1}A sample of 4 features shown to illustrate cluster membership\\
\end{minipage}

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562
days) with median survival time calculated as 150 days. At time 0 the
survival probability was 0.82 (95\% CI: 0.69, 0.98). Survival
probability after one year of 0.403 (95\% CI: 0.252, 0.644) and survival
probability at close to ten years of 0.24 (95\% CI: 0.11, 0.52).

@@

{\centering \includegraphics{paper_files/figure-pdf/fig-box-1.pdf}

\caption{\label{fig-box}A compariative box plot showing the range of
survival times for subjects split by the type of variant. Green bars
indicate homozygous variants, while red indicates compound heterozygous
variants.}

@@

This meta-review of current research was able to consolidate data that
allowed statistical analysis to be conducted on affected individuals.
The use of dimension reduction techniques was shown to be successful in
uncovering relevant and novel substructure in the data. This was
particularly evident with the detection of the accepted typical
phenotype of microcephaly, hypomyelination, arthrogryposis, simplified
gyral pattern and developmental delay. After excluding two outlying
remarkable cases, we identified the presence of an adjacent cluster of
related features that are not well described in The Human Phenotype
Ontology \citep{kohler2021human} as shown in Table~\ref{tbl-hpo}. It
shows the prolific feature of `respiratory distress' as more commonly
associated with these adjacent features such as vocal cord palsy,
tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD).
These features are present in the case reported data, but not well
described in HPO and help account for the variation seen in the clinical
phenotype in the first few principal dimensions.

The heterogeneous variation in the data was clearly evident with the
first four principal axes being required to explain more than 50\% of
the variation in the data set. Visual inspection shows the primary
sources of variation in the clinical phenotype are explained by a few
individuals with very different clinical features. Many of these exhibit
highly specific facial or anatomical dysmorphisms. While coarse facial
features are a core presentation, those with a milder course had
clinical reports that articulated these dysmorphisms in much more
detail. Additionally, talipes appears as an isolated feature without the
typical microcephaly and hypomyelination, however there was significant
heterogeneity in how feet contractures were reported in clinical case
reports. This is particularly evident in rocker-bottom feet which was
found closely related in our analysis to core features despite the
prevalence in HPO being quite low (2/21), suggesting possible
under-reporting or confounded reporting of this feature. Another core
identified feature is bilateral cleft lip. This condition has a reported
frequency of just 2/21 (Table~\ref{tbl-hpo}) and may be an emerging
feature for a more typical presentation given its close relationship
with other classic features. Features isolated to just one or few cases
in our consolidated data have been labelled as `Isolated' in
Table~\ref{tbl-hpo}. An optimal choice of both 2 and 5 clusters were
recommended in the analysis. A partition of the data with just two
clusters would likely be useful in separating outliers from other data.
In this analysis a larger number (5) was selected to align the goals of
the analysis which was to uncover distinct groups of features.

In terms of patient survival, the first few months of life show a sharp
decline in survival probability confirming early demise in the infant
period as a key factor. A small number of cases show survival into
childhood and in some cases into the second decade of life. This work
identifies a one year survival rate of 0.403 (95\% CI: 0.252, 0.644). On
the other hand, a naive estimate using just the number of deceased
patients at this time as proportion of the study population is 0.43 (1 -
16/28). This highlights the importance of applying a survival analysis
technique that incorporates censoring so as not to erroneously
overestimate the survival probability. The current literature suggests
the type of gene variant has an impact on survival time, which seems to
be supported in this work. Through a comparison of multiple survival
curves, stratified by variation type, the limited sample size prevented
definitive conclusions with many cells having expected counts less than
5.
